How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar

### 2020-02-18 docs/manuscript.Rmd
updating organization of man

---
title: |
 How accesss to care shapes disease burden: the current impact of post-exposure prophylaxis 
 and potential for expanded access to prevent human rabies deaths in Madagascar
author:
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [Princeton]
      correspondence: true
  - Helene Guis:
      institute: [IPMepi, CIRAD]
      correspondence: false
  - Glenn Edosoa:
      institute: [OMS]
      correspondence: false
  - Chantal Hanitriniaina:
      institute: [UoA]
      correspondence: false
  - Soa Fy Andriamandimby:
      institute: [IPMviro]
      correspondence: false
  - Anjasoa:
      institute: [IPMepi]
      correspondence: false
  - Reziky:
      institute: [IPMepi]
      correspondence: false
  - Ravo:
      institute: [IPMvacc]
      correspondence: false
  - Laurence Baril:
      institute: [IC]
      correspondence: false
  - C.J.E. Metcalf:
      institute: [Princeton]
      correspondence: false
  - Katie Hampson:
      institute: [Glasgow]
      correspondence: false
institute:
  - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IC: Institut Pasteur de Cambodge
  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
  - OMS: OMS
  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
date: ""
link-citations: no
output:
 bookdown::word_document2:
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
      - --reference-doc=format/Reference_Document.docx
 bookdown::gitbook:
    toc: true
    toc_float: true
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
bibliography: ["bib/refs_all.bib"]
---

<!-- To Do: -->
<!-- 1) Go through methods + results -->
<!--   a) update figs -->
<!--   b) update references -->
<!--   c) add in stats -->
<!-- 2) Work through comments in discussion per KH -->
<!-- 3) Insert citations for discussion -->
<!-- 4) Work through introductions -->
<!-- 5) Think through doc workflow (how to split docs & refs) -->

```{r setup, include=FALSE}
## Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(rgdal)

opts_chunk$set(echo = FALSE)
out_main_fig = TRUE

@@

# ABSTRACT

@@

# INTRODUCTION
[see figure 1 A](#fig1A)

# METHODS
## Spatial covariate data {#methods-spatial}
To calculate travel times, we used the global friction surface for 2015 generated by the Malaria Atlas Project (https://map.ox.ac.uk/research-project/accessibility_to_cities/, @weissGlobalMapTravel2018) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at an approximately 1 x 1 km scale. We calculated a weighted average of travel times by human population at the commune and district levels, using administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs ([see section S1](#S1)). Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, @linardPopulationDistributionSettlement2012) and aggregated to the commune and district levels.

## Bite patient data {#methods-bitedata}
We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted from clinics as frequently as monthly to annually. These data include details of the administrative district of the bite patient’s address and the date of reporting.

```{r fig 1, fig.fullwidth=TRUE}
include_graphics("../figs/main/M1.jpeg")
```

*[__Figure 1.__]{#fig1} __A)__ The network of patient presentations at the district level for the national data: circles with black outline represent the total number of patients reporting to each clinic for which we have data. Color corresponds to the clinic catchment. Circles with white outline are the total number of bites reported for that district, and the lines show which ARMC those patients reported to, with the line width proportional to number of patients. Districts in catchments excluded due to lack of forms submitted by the clinic are colored in grey. __B)__ The network of patient presentations at the commune level for the Moramanga data with inset of Madagascar showing the location of the district. Circle with black outline is the location of the Moramanga ARMC with the size proportional to the total number of bites reported to the clinic. __C)__ The estimated average annual bite incidence corrected for reporting for the years 2014 - 2017 (y-axis) for each district plotted against travel time to the closest ARMC (x-axis). Colors correspond to the clinic catchment, 
shape indicates the dataset, and the size of the point indicates the number of observations (i.e. the number of years for which data was available for the national data; note for Moramanga `r max(mora_bites$nobs, na.rm = TRUE)` months of data were used).*

For clinics that did submit data, there was substantial underreporting, with months to even a whole year when forms were not submitted. We therefore estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days (Figure S2.1A and B). To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level. We found that these corrected estimates of reported bites was in agreement with the vial demand for clinics (Fig S2.1C, [further details in section S2](#S2)).

```{r contacts stats}
contacts_mora <- fread("../output/sensitivity/contacts_mora.csv")
contacts_natl <- fread("../output/sensitivity/contacts_natl.csv")
prop_excluded_natl <- (sum(contacts_natl[estimated_cat1 == 1]$no_patients) + sum(contacts_natl[estimated_cat1 == 0 & !is.na(no_known)]$no_known))/sum(contacts_natl$no_patients)

```

Previous work in the Moramanga District showed that Category I exposures, i.e. touching or feeding a suspected animal or human rabies case, make up approximately 20% of patients reporting to ARMCs [@rajeevHealthcareUtilizationProvisioning2018]. These low risk exposures do not contribute to the mortality burden of rabies. However, we could not directly exclude them as details on exposure type was missing for the majority of the patient forms. Detailed data from Moramanga indicates that these contacts consistently present as clustered cases and the majority report on a single day. Accordingly, we excluded patients reporting on any dates that had greater than 3 standard deviations above the mean number of patients reporting per day (Figure S2.2A). We validated this method using the Moramanga ARMC data for which we had details on the type of exposure, and found that setting the threshold to 3 standard deviations (SDs) excluded approximately `r round(contacts_mora[prop == "contacts" & sd == 3.00]$excluded*100, 2)`% of Category I exposures, with only `r round(contacts_mora[prop == "bites" & sd == 3.00]$excluded*100, 2)`% of Category II and III patients excluded (Figure S2.2B). We further excluded patients which were explicitly noted to be Categort I exposures but were not identified based on clustering of presentation dates, resulting in the exclusion of approximately `r round(prop_excluded_natl*100, 2)`% of patient records from the national data. After excluding patients likely to reflect Category I exposures as described, we estimated average bite incidence as the mean of the yearly bites reported/the proportion of bite patient data which were submitted.

The 31 months of data available from the Moramanga district resolved to the commune level (the administrative level below the district, Figure 1B) provide a useful comparison to the national scale data. These data were completely reported with details on the type of exposure documented. We calculated average annual bite incidence to the commune level by excluding Category I exposures and taking the monthly average of reported bites and multiplying by 12. Our final dataset consisted of estimates of average bite incidence for `r nrow(district_bites)` out for `r length(mada_districts)` districts total, and `r nrow(mora_bites)` communes within the catchment of the Moramanga District (Fig 1C).

## Model of reported bites as a function of travel time to closest ARMC {#methods-bitemods}
We modeled incidence of bites as a function of travel time ($T$) as follows: 
$$ \mu = exp(\beta_{t}T \ + \ \beta_0)P$$
where $\mu$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC, and the human population size, $P$ (an offset which scales the predicted number of bites), for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian framework via MCMC using the R package ‘rjags’. In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:

@@

## Estimating burden of rabies deaths {#methods-burden}
We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 2). We assume annual rabies incidence of 1% in the dog population in an endemic setting with no mass dog vaccination [@consortiumPotentialImpactImproved2019]. To generate $E_i$, the human exposure incidence in a given administrative unit (where $i$ is the index of the commune or district) based on the model, we multiply this baseline incidence by the estimated dog population and the exposure rate per rabid dog ($p_{exp}$). We estimate the dog population given a human:dog ratio (HDR) of 7 to generate our maximum expected incidence and an HDR of 35 for our minimum expected incidence, giving a range of 11.1 ($E_{min}$) - 55.7 ($E_{max}$) exposures per 100,000 persons. As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range encompasses both a wide range of observed HDRs and human exposure incidence across Africa [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating]. $E_i$ is then drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location.

*__Figure 2.__Decision tree for burden estimation. For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$). To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$). $E_i$ is generated from assumptions about rabies exposure incidence and the underlying dog population extrapolated from human population sizes (see Table 1, Methods text). $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$). Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e. those who would have died given exposure, but instead received PEP. All parameter values and sources are listed in Table 1.*

@@

| **Parameter**   | **Value**        | **Description**	                 | **Source**|
|:----------------|:----------------:|:----------------------------------|:---------:|
| $p_{exp}$ | 0.39 | Probability of human exposure given a dog rabies case| [@consortiumPotentialImpactImproved2019] |
| $E_{min} − E_{max}$ |	11.1 - 55.7 per 100,000 persons	| Range of human exposure incidence calculated from an assumed dog rabies incidence of 1%, human:dog ratios between 7 - 35, and $p_{exp}$. | [@consortiumPotentialImpactImproved2019; @knobelReevaluatingBurdenRabies2005; @sambo2018estimating]|
| $p_{rabid}$	| 0.2 - 0.6 |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
| $\rho_{max}$ | 0.98	| The maximum reporting possible for any location	Data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2018] |
| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|

## Strategically expanding PEP access {#methods-incremental}

@@

We developed a framework to rank clinics by how much they could improve access for people currently underserved (< 3 hours away from any clinic in the country) and to estimate the incremental reduction in burden as PEP is expanded to these clinics. We use the approximate locations of all existing public CSB II (centre de sante niveau II, the level below district hospitals, n = `r nrow(csbs)`) as locations of potential ARMC.

Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country. We added the clinic which minimized this metric weighted by the reduction in travel times, and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially. In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons). We also estimate travel times given that all CSB II in the country provision PEP to get an estimate of the maximum impact of expanding PEP provisioning. Finally, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.

We also estimate how expansion to additional clinics may impact vaccine demand. Specifically, we use our model-predicted estimates of annual bites and then aggregate these to the clinic catchment. As clinics are added, multiple clinics may serve a single district, but in general, the majority of the population within a given commune are closest to a single clinic (Fig S5.3B). Therefore, we assign the catchment clinic as the one which is the closest for the majority of people living in a commune. For the district model, we disaggregate the reported bites to the commune level (assuming uniform incidence within the district). For each catchment, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e. all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses. We use these dates to estimate vial usage given routine vial sharing practices in Madagascar (i.e. each vial is split between two patients, and vials are discarded at the end of each day) and the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]). We assume that vial sharing is functioning perfectly, with a background wastage of 0.1 mL per vial, but not accounting for wastage due to other reasons (i.e. wastage due to injection error). We take the mean of 100 simulations of throughput for each catchment and set of clinics.

## Sensitivity analyses for burden {#methods-sensitivity}
We estimate burden across a range of parameter values to test the effects of our model assumptions on estimates of rabies burden. Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, $p_{rabid}$ at 0.2 and 0.6, and $\rho_{max}$ at 0.85 and 0.99 to obtain maximum and minimum estimates of burden. We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population, see [supplement section S6](#S6) for more details) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size.

@@

## Travel times as a predictor of bites {#results-bitemods}
All of the candidate models which included travel times produced reasonable predictions when compared to the observed data (Figure SY). Population size alone was the poorest fit to the data as estimated by DIC, and all models with population size as a covariate did not generate realistic predictions to the data or when used to predict out of fit. For the national data, the models with travel times alone as a covariate and a random catchment generated predictions closest to the observed data (Figure 3B, Figure SX). In addition, the estimates from the model fitted to the Moramanga data fall into the prediction envelope for the random catchment intercept models (Figure 3A). As these models likely capture the range of possible relationships for any given catchment in our data set, we use estimates from these models to generate subsequent predictions of bite incidence and burden. 
_**Figure 3.** Travel times as a predictor of reported bite incidence per 100,000 persons A) the estimated relationship between travel times in hours (x-axis) and mean annual bite incidence per 100,000 persons (y-axis). The lines are the estimates using the mean of the posterior distributions of the parameters for three candidate models: model with travel times at the 1) commune and 2) district level fitted to the national data with a random intercept by catchment, and 3) travel times at the commune level fitted to the Moramanga data with a fixed intercept. The envelopes shows the 95% CI given the random intercept for the commune and district models. The Moramanga data was fitted with a fixed intercept, but falls in the prediction envelope for the commune model with a random intercept.  B) Log(observed) vs. log(predicted) mean annual reported bites for the three different models (added a small positive offset of 0.1 to account for log of zeros)._

>>>>

---
title: "How access to care shapes disease burden: the current impact of post-exposure prophylaxis and potential for expanded access to prevent human rabies deaths in Madagascar"
author:
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [Princeton]
      correspondence: true
  - Helene Guis:
      institute: [IPMepi, CIRAD]
      correspondence: false
  - Glenn Edosoa:
      institute: [OMS]
      correspondence: false
  - Chantal Hanitriniaina:
      institute: [UoA]
      correspondence: false
  - Soa Fy Andriamandimby:
      institute: [IPMviro]
      correspondence: false
  - Anjasoa Randrianarijaona:
      institute: [IPMepi]
      correspondence: false
  - Reziky Tiandraza Mangahasimbola:
      institute: [IPMepi]
      correspondence: false
  - Ravo Ramiandrasoa:
      institute: [IPMvacc]
      correspondence: false
  - Laurence Baril:
      institute: [IC]
      correspondence: false
  - C.J.E. Metcalf:
      institute: [Princeton]
      correspondence: false
  - Katie Hampson:
      institute: [Glasgow]
      correspondence: false
institute:
  - Princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - IPMviro: Virology Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMvacc: Vaccination Center, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IPMepi: Epidemiology and Clinical Research Unit, Institut Pasteur de Madagascar, Antananarivo, Madagascar
  - IC: Institut Pasteur de Cambodge
  - CIRAD: CIRAD, UMR ASTRE, Antananarivo, Madagascar
  - OMS: OMS
  - UoA: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - Glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
date: ""
link-citations: no
output:
 bookdown::word_document2:
    pandoc_args:
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
      - --reference-doc=format/Reference_Document.docx
 bookdown::html_document2:
    toc: true
    toc_float: true
    pandoc_args:
      - --csl=format/plos-neglected-tropical-diseases.csl
      - --filter=pandoc-crossref
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
      - --lua-filter=format/multiple-bibliographies.lua
keywords: "canine rabies, post-exposure prophylaxis, Gavi, Zero by 30, access to care"
bibliography: ["bib/refs_all.bib"]
# bibliography_main: ["bib/refs_main.bib"]
# bibliography_supp: ["bib/refs_supp.bib"]
---

##### TO DO: fix affiliations

```{r setupmain, include=FALSE}
## Libraries
library(knitr)
library(tidyverse)
library(data.table)
library(rgdal)
library(glue)

opts_chunk$set(echo = FALSE, warning = FALSE)
out_main_fig = TRUE

@@

# ABSTRACT
Access to care shapes burden of diseases globally. For dog-mediated rabies, post-exposure prophylaxis (PEP) is highly effective at preventing human deaths, however access to PEP is limited in rabies endemic countries. Recently, a target to eliminate deaths due to canine rabies has been set forth by global health partners. In 2018, Gavi added human rabies vaccine to its investment portfolio, a decision which has the potential to greatly reduce rabies deaths. Here, we estimate rabies burden in Madagascar incorporating estimates of geographic access to PEP. We find that travel times to the closest Anti-Rabies Medical Center (ARMC) is a strong predictor of reported bite incidence across the country. We use resulting estimates of bite incidence in an adapted decision tree framework to generate estimates of burden and reporting across the country, and find that spatial access to PEP shapes local burden. We evaluate the impact of expanding access to PEP, using a strategic framework which minimizes travel time for people at the greatest risk of death given a rabies exposure (persons > 3 hrs away from a clinic). Under these assumptions, we estimate that provisioning of PEP to an additional 200 clinics in Madagascar could reduce the burden of rabies deaths by 50%.  These results highlight the importance of accounting for access to care in predicting the impacts of interventions sub-nationally and the potential for expanding PEP access strategically to reduce rabies deaths in endemic settings.

@@

# INTRODUCTION

Inequities in access to care are a major driver of disease burden globally [@world2018world]. Often, the populations at greatest risk of a given disease are the most underserved [@peters2008poverty]. Delivering interventions to these groups is challenging due to financial and infrastructural limitations, and requires careful consideration of how best to allocate limited resources [@mainaSpatialDatabaseHealth2019].

Dog-mediated rabies is estimated to cause approximately 60,000 human deaths annually [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. Mass vaccination of domestic dogs has been demonstrated to be a highly effective way to control the disease in both animals and humans . While dog vaccination can interrput transmission in the reservoir, human deaths can also be prevented through prompt administration of post-exposure prophylactic vaccines (PEP) following a bite by a rabid dog [@RabiesVaccinesWHO2018]. However, access to human vaccines are limited in countries where canine rabies is endemic [@sreenivasanOverviewRabiesPostexposure2019; @tarantolaRabiesPostexposureProphylaxis2017; @hampsonEstimatingGlobalBurden2015], and within countries these deaths are often concentrated in rural, underserved communities [@wentworthSocialJusticePerspective2019].

In 2015, a global framework to eliminate deaths due to dog-mediated rabies by 2030 through a combination of PEP provisioning and dog vaccination was established by the World Health Organization (WHO) and partners [@world2018zero]. Furthermore, in 2018, Gavi, the Vaccine Alliance, added human rabies vaccines to their proposed investment portfolio [@GaviVIS]. From 2021, Gavi-eligible countries will be able to apply for support to expand access to these vaccines, with potential to greatly reduce deaths due to rabies. Yet it is unclear how these vaccines should be distributed to prevent deaths most effectively.

A primary challenge is the lack of data on rabies exposures in humans and incidence in animals in most rabies-endemic countries [@scottAddressingDisconnectEstimated2017]. The majority of rabies burden studies at the national level have integrated bite patient data on reported bites into a probability decision tree framework, with the key assumption that overall reported bite incidence (i.e. both bites due to non-rabid and rabid animals) are proportional to rabies incidence (i.e. the more bites reported in a location, the higher the incidence of *rabies* exposures), and that reporting is uniform across space [@hampsonEstimatingGlobalBurden2015; @knobelReevaluatingBurdenRabies2005; @rajeevHealthcareUtilizationProvisioning2018]. If applied subnationally, these assumptions would likely underestimate rabies deaths in places with poor access to PEP and may overestimate rabies deaths in places with better access to PEP.

In Madagascar, the Institut Pasteur de Madagascar (IPM) provides PEP at no-cost to patients at 31 clinics, or anti-rabies medical centers (ARMC) across the country [@rajeevHealthcareUtilizationProvisioning2018]. PEP is not available at any other public clinics or through the private sector. In addition, there is limited control of rabies in dog populations and the disease appears to be endemic throughout the country [@andriamandimbySurveillanceControlRabies2013; @reynesLaboratorySurveillanceRabies2011]. Due to the spatially restricted nature of PEP distribution points and lack of direct costs for PEP, geographic access is likely to be a major driver of disease burden within the country. Previously, we  estimated the burden of rabies in Madagascar nationally using data from a single district to extrapolate to the country, and did not account for spatial variation in access [@rajeevHealthcareUtilizationProvisioning2018]. Here, we revise estimates of burden by incorporating the impact of access to PEP at the sub-national level on preventing human rabies deaths and explore the potential impact of expanding provisioning of human rabies vaccines on further reducing these deaths.

# METHODS

## Estimating travel time to closest ARMC {#methods-spatial}
We used the global friction surface for 2015 generated by the Malaria Atlas Project (https://map.ox.ac.uk/research-project/accessibility_to_cities/, @weissGlobalMapTravel2018) and GPS points of clinics to estimate the travel time to the nearest ARMC for the country of Madagascar at an approximately 1 x 1 km scale. We calculated a weighted average of travel times by human population at the commune and district levels, using administrative shapefiles from the UN Office for the Coordination of Humanitarian Affairs ([see supplementary section S1](#S1)). To asses the accuracy of this method, we compared this estimate to driving time data collected by IPM during field missions and self-reported travel times from a subset of patients reporting to the Moramanga ARMC. We also compared unweighted average travel times and distance  Human population estimates were taken from the 2015 UN adjusted population projections from World Pop (www.worldpop.org, [cite older WP paper here!]) and aggregated to the commune and district levels. We compared these population estimates to recently released estimates from the 2018 census of Madagascar, and a recent census of three communes in the Moramanga District (see SZ).

## Estimating average annual bite incidence {#methods-bitedata}
We used a database of bite patient forms submitted to IPM from ARMC across the country between 2014 - 2017. These were individual patient data forms that were submitted from clinics as frequently as monthly to annually. These data include details of the administrative district of the bite patient’s address and the date of reporting.

For clinics that did submit data, there was substantial underreporting, with months to even a whole year when forms were not submitted. We therefore estimated clinic level reporting as the proportion of days on which forms were submitted, excluding any periods for which there were no forms submitted for 15 consecutive days ([Figure S2.1A and B](#figS2.1)). To estimate the average annual bites reported for each district, we further excluded data from any years for which there was less than 25% reporting at the clinic level.

The 31 months of data available from the Moramanga district resolved to the commune level (the administrative level below the district, Figure 1B) provide a useful comparison to the national scale data. These data were completely reported with details on the type of exposure documented. We calculated average annual bite incidence to the commune level by taking the monthly average of reported bites and multiplying by 12.

## Modeling reported bite incidence as a function of travel time to closest ARMC {#methods-bitemods}
We modeled incidence of bites as a function of travel time ($T$) as follows: 
$$ \mu = exp(\beta_{t}T \ + \ \beta_0)P$$
where $\mu$ is the expected number of bite patients presenting at the ARMC as a function of travel time to the ARMC, and the human population size, $P$ (an offset which scales the predicted number of bites), for a given source location (district or commune). We fit this model to both the national data resolved to the district level and the Moramanga data resolved to the commune level in a Bayesian framework via MCMC using the R package ‘rjags’. In order to more directly compare estimates between these two datasets, we also modeled the national data with a commune-level travel time covariate ($T_{j}$) so that:

@@

Finally, to check how correcting for underreporting affected our modeling results, we fitted these same models to data uncorrected for reporting (see [supplementary section S4](#S4)).

## Modeling the burden of human rabies deaths sub-nationally {#methods-burden}

<!-- TO DO: simplify discussion here but bring in some figs, etc to Supplementary section to explain assumptions! -->

We estimate deaths due to rabies as a function of the number of bites predicted by our model and estimates of endemic rabies exposure incidence using an adapted decision tree framework (Fig 2). We assume annual rabies incidence of 1% in the dog population in an endemic setting with no mass dog vaccination [@whorabiesmodelingconsortiumPotentialImpactImproved2019]. To generate $E_i$, the human exposure incidence in a given administrative unit (where $i$ is the index of the commune or district) based on the model, we multiply this baseline incidence by the estimated dog population and the exposure rate per rabid dog ($p_{exp}$). We estimate the dog population given a human:dog ratio (HDR) of 7 to generate our maximum expected incidence and an HDR of 35 for our minimum expected incidence, giving a range of 11.1 ($E_{min}$) - 55.7 ($E_{max}$) exposures per 100,000 persons. As there is little data on dog populations in Madagascar [@andriamandimbySurveillanceControlRabies2013], this range encompasses both a wide range of observed HDRs [@knobelReevaluatingBurdenRabies2005; @sambo2018estimating] and generates similar human exposure incidences as observed previously across Africa [cite here!] . $E_i$ is then drawn from a uniform distribution between the minimum and maximum expected number of human exposures for each location.

```{r fig 2, fig.fullwidth=TRUE}
include_graphics("../figs/main/M2.jpeg")
```
*__Figure 2.__Decision tree for burden estimation. For a given administrative unit $i$, human deaths due to rabies ($D_i$) are calculated from model predicted bites ($B_i$). To get $R_i$, the number of reported bites that were rabies exposures, we multiply $B_i$ by $p_{rabid}$, the proportion of reported bites that are rabies exposures. $p_{rabid}$ is constrained such that reported rabies exposures cannot exceed total rabies exposures ($E_i$) and reporting cannot exceed a maximum ($p_{max}$). $E_i$ is generated from assumptions about rabies exposure incidence and the underlying dog population extrapolated from human population sizes (see Table 1, Methods text). $R_i$ is subtracted from $E_i$ to get the number of unreported bites ($U_i$) and then multiplied by the probability of death given a rabies exposure ($p_{infect}$) to get the number of deaths due to rabies ($D_i$). Similarly, deaths averted by PEP are estimated by multiplying $R_i$ by $p_{infect}$, i.e. those who would have died given exposure, but instead received PEP. All parameter values and sources are listed in Table 1.*

@@

| **Parameter**   | **Value**        | **Description**	                 | **Source**|
|:----------------|:----------------:|:----------------------------------|:---------:|
| $p_{exp}$ | 0.39 | Probability of human exposure given a dog rabies case| [@whorabiesmodelingconsortiumPotentialImpactImproved2019] |
| $E_{min} − E_{max}$ |	11.1 - 55.7 per 100,000 persons	| Range of human exposure incidence calculated from an assumed dog rabies incidence of 1%, human:dog ratios between 7 - 35, and $p_{exp}$. | [@whorabiesmodelingconsortiumPotentialImpactImproved2019; @knobelReevaluatingBurdenRabies2005; @sambo2018estimating]|
| $p_{rabid}$	| 0.2 - 0.6 |	Proportion of reported bites that are rabies exposures | [@rajeevHealthcareUtilizationProvisioning2018] |
| $\rho_{max}$ | 0.98	| The maximum reporting possible for any location	Data from Moramanga Ville, Moramanga District | [@rajeevHealthcareUtilizationProvisioning2018] |
| $p_{death}$ |	0.16 | The probability of death given a rabies exposure	| [@changaluchaNeedImproveAccess2018]|

## Estimating the impact of expanding PEP provisioning to additional clinics {#methods-incremental}

@@

We developed a framework to rank clinics by how much they could improve access for people currently underserved (> 3 hours away from any clinic in the country) and to estimate the incremental reduction in burden as PEP is expanded to these clinics. We use the approximate locations of all existing public CSB II (centre de sante niveau II, the level below district hospitals, n = `r nrow(csbs)`) as locations of potential ARMC.

Starting with the current locations, we added one clinic at a time, calculating the proportion of people living < 3 hours away from any clinic for the country. We added the clinic which minimized this metric weighted by the reduction in travel times, and then repeated the process iteratively, ranking clinics and adding the top clinic sequentially. In this iterative process, we also exclude any clinics that once added only reduced travel times below the 3 hour threshold for less than 0.01% of the population (< `r round(sum(mada_districts$pop)*0.0001, -2)` persons). We also explore the scenario in which all CSB II in the country provision PEP to get an estimate of the maximum impact of expanding PEP provisioning. Resulting travel time estimates and locations of when and where clinics were added are shown in [Supplementary section S5](#S5). Finally, we estimate the incremental reduction in burden as each clinic is added using the burden framework described in the previous section.

We estimate how expansion to additional clinics may impact vaccine demand. Specifically, we use our model-predicted estimates of annual bites and then aggregate these to the clinic catchment. As clinics are added, multiple clinics may serve a single district, but in general, the majority of the population within a given commune are closest to a single clinic (Fig S5.3B). Therefore, we assign the catchment clinic as the one which is the closest for the majority of people living in a commune. For the district model, we disaggregate the reported bites to the commune level (assuming uniform incidence within the district). For each catchment, we simulate throughput by randomly assigning patient presentation dates, and then assume perfect compliance (i.e. all patients reporting 3 and 7 days after their first dose) to generate dates of subsequent doses. We use these dates to estimate vial usage given routine vial sharing practices in Madagascar (i.e. each vial is split between two patients, and vials are discarded at the end of each day) and the new abridged intradermal regimen recommended by WHO (2 injection sites on day 0, 3, and 7, [@tarantolaIntradermalRabiesPostexposure2018]). We assume that vial sharing is functioning perfectly, with a background wastage of 0.1 mL per vial, but not accounting for wastage due to other reasons (i.e. wastage due to injection error). We take the mean of 1000 simulations of throughput for each catchment and set of clinics.

## Sensitivity of burden estimates to decision tree parameters {#methods-sensitivity}
We estimate burden across a range of parameter values to test the effects of our model assumptions on estimates of rabies burden. Specifically, we fix rabies incidence at the minimum and maximum of our estimated range, $p_{rabid}$ at 0.2 and 0.6, and $\rho_{max}$ at 0.85 and 0.99 to obtain maximum and minimum estimates of burden. We also examine the impact of systematic variation in rabies incidence with human population (a potential proxy for changes in the dog population, see [supplementary section S6](#S6) for more details) by looking at how estimates of burden change if rabies incidence were to scale positively or negatively with human population size.

## Data availability
All analyses were done in R and using these packages (list packages with commas and citations). All processed data, code, and outputs are archived at this zenodo repository and maintained at this github repo. The raw bite patient data at the national level is maintained by IPM. This database was last queried on *date here* for these analyses. Anonymized raw data from this database are available upon request from IPM. Anonymized raw data collected from the Moramanga District are shared in the archived repository.

## Ethics statement
Data collection from the Moramanga District was approved by the Princeton University IRB (# 7801) and the Madagascar Ministry of Public Health Ethics Committee (# 105-MSANP/CE). Oral informed consent was obtained from all interviewed participants. Data collected from bite patients at the national level are maintained jointly by the Ministry of Health and IPM as a routine part of PEP provisioning.

@@

## Estimating travel times to and catchment populations of 
Travel times to the closest ARMC varies greatly across the country. We compared these estimates based
on the MAP friction surface to self-reported estimates of hours traveled to the Moramanga clinic for a subset of patients and driving times across the country from IPM. We found that the travel times estimated from the friction surface underestimate true driving times and patient reported travel times to the Moramanga ARMC. Importantly, the friction surface does not account for road condition, traffic, or seasonality in travel times. It also assumes that given multiple types of transport in a cell, the fastest mode of transport takes precedence. From the self reported travel patterns, we know that people use mutliple modes of transport and walking is common adding to the variation (SX). Despite these limitations, the estimated travel times correlate strongly with the reported and groundtruthed driving times. Travel times weighted by population at the grid cell level were a better predicter than unweighted travel times or distance (SY). We use the estimated weighted mean of travel times by population as a proxy for access at the commune/district level in subsequent analyses.
